To Batch or Not to Batch: Sequential vs. Batched Testing Strategies in the ED

### 2023-11-09 manuscript/manuscript.Rmd
breaktthough on the manuscript interpretation

In this section, we examine the causal effects of provider batching tendency. One thing that we note in the

We first present the effects on our main outcomes of length of stay, 72hr readmission, and healthcare resource utilization (number of tests). We find statistically and economically significant positive effects of assignment to a high batching physician on every outcome (Table 3). Estimated coefficients are scaled by the difference in leniency going from the ninetieth to tenth percentile in physician leniency—equal to 11.6 pp—for interpretability. Assignment to a physician in the top prescribing decile (relative to one in the bottom decile) is associated with a 0.24 pp increase in the probability of long-term prescription opioid use— a 4.1 percent increase. The effect on proxies for opioid-seeking behavior is 0.50 pp—a 3.4 percent increase. Effects on development of opioid use disorder and on opioid overdose mortality are 2.1 percent and 8.8 percent, respec- tively. These findings highlight the substantial role ED physicians can play in putting patients on a path of long-term opioid use and dependence.

\begin{table}[htbp]
\centering
\caption{Reduced Form Regression Results}
\label{tab:reducedform}
\begin{tabular}{p{6cm}ccc}
\toprule
& \multicolumn{1}{c}{\textbf{Log ED LOS}}
& \multicolumn{1}{c}{\textbf{Number of Tests}}
& \multicolumn{1}{c}{\textbf{72hr Return}} \\
& \multicolumn{1}{c}{(1)}
& \multicolumn{1}{c}{(2)}
& \multicolumn{1}{c}{(3)} \\
\midrule
\textit{Model 1} & & & \\
Batch Tendency      & $0.905^{***}$ & $1.715^{***}$ & $-0.025$  \\
                    & $(0.308)$  & $(0.329)$     & $(0.019)$ \\
\\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Patient observables? & Yes & Yes & Yes \\
\midrule
\midrule
\textit{Model 2} & & & \\
Batch Tendency      & $-0.103$      & $0.120^{***}$ & $0.020$   \\
                    & $(0.462)$     & $(0.046)$     & $(0.021)$ \\
                    &     &     &  \\
Testing Inclination    & $0.608^{**}$  & $0.963^{***}$ & $-0.027^{**}$ \\
                    & $(0.242)$  & $(0.021)$     & $(0.011)$ \\
\\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Patient observables? & Yes & Yes & Yes \\
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{Estimates of the reduced form are for the baseline sample described in the text. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

>>>>

In this section, we examine the causal effects of provider batching tendency. Our reduced form regression results, as presented in Table \ref{tab:reducedform}, indicate that the physician's batch tendency is associated with longer emergency department length of stay (ED LOS), a greater number of tests ordered, and does not significantly influence the likelihood of a 72-hour return. However, the relationship between batch tendency and our outcomes of interest is nuanced by the presence of testing inclination among physicians.

\begin{table}[htbp]
\centering
\caption{Reduced Form Regression Results with Scaled Coefficients}
\label{tab:reducedform}
\begin{tabular}{p{6cm}ccc}
\toprule
& \multicolumn{1}{c}{\textbf{Log ED LOS}}
& \multicolumn{1}{c}{\textbf{Number of Tests}}
& \multicolumn{1}{c}{\textbf{72hr Return}} \\
& \multicolumn{1}{c}{(1)}
& \multicolumn{1}{c}{(2)}
& \multicolumn{1}{c}{(3)} \\
\midrule
\textit{Model 1} & & & \\
Batch Tendency      & $0.138^{***}$ & $0.270^{***}$ & $-0.004$  \\
                    & $(0.047)$  & $(0.052)$     & $(0.003)$ \\
\\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Patient observables? & Yes & Yes & Yes \\
Hospital volume? & Yes & Yes & Yes \\
\midrule
\midrule
\textit{Model 2} & & & \\
Batch Tendency      & $-0.018$      & $0.019^{**}$ & $0.003$   \\
                    & $(0.071)$     & $(0.007)$     & $(0.003)$ \\
                    &     &     &  \\
Testing Inclination    & $0.595^{**}$  & $0.962^{***}$ & $-0.027^{**}$ \\
                    & $(0.234)$  & $(0.021)$     & $(0.011)$ \\
\\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Patient observables? & Yes & Yes & Yes \\
Hospital volume? & Yes & Yes & Yes \\
\midrule
Mean number of encounters per physician & 1805.3 &  &  \\
\midrule
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{Estimates of the reduced form are for the baseline sample described in the text. Coefficients on batch tendency and testing inclination are scaled by the difference between batch tendency and testing inclination respectively of the ninetieth and tenth percentile physicians for interpretability. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

Similar to our method used to measure physician batch tendency, we use the physician's residualized leave-out average test rate to get a measure of a physicians propensity to test (which we call testing inclination). A key consideration in our analysis is the strong positive correlation (r = 0.79) between batch tendency and testing inclination. Such a high degree of correlation raises the concern that physicians with a higher testing inclination may batch more often because as a result of ordering more tests. Failure to control for this inclination could result in biased estimates of the effect of batch tendency. We note this assumption made on the causal pathway and further explore this in the section on mediation analysis.

In Model 2, we control for testing inclination and observe notable changes in the coefficients for batch tendency. The previously significant positive association with Log ED LOS diminishes, suggesting that the initial estimate may have been capturing the combined effects of batching and testing inclination. Estimated coefficients for batch tendency are scaled by the difference in batch tendency going from the ninetieth to tenth percentile in physician batch tendency—equal to 15.8 pp—for interpretability, and coefficients for testing inclination are scaled by the difference in testing inclination going from the ninetieth to tenth percentile in physician testing inclination—equal to 34.9 pp.

Assignment to a physician in the top batching decile (relative to one in the bottom decile) is associated with a 0.2 unit increase in the number of diagnostic tests being ordered per patient encounter, without any additional change in the liklihood of a 72 hour return. With an average of 1,805 patient encounters per physician in our study period, this translates to an additional 8,667 diagnostic tests with no perceived benefit in terms of reducing returns or length of stay.

The effects of testing inclination are also striking. We find that a movement from a physician in the top testing decile (relative to one in the bottom decile) is associated with 0.6% change in the length of stay in the emergency department, a 0.96 unit increase in the number of diagnostic tests being ordered per patient encounter, and -0.03 pp change in the probability of a patient returning within 72 hours. Given the average number of patient encounters per physician in our study period this translates to an additional 1088 patient hours in LOS, 41,594 diagnostic tests, and 47 avoided 72 hour returns.

### 2023-11-09 manuscript/manuscript.tex
breaktthough on the manuscript interpretation

In this section, we examine the causal effects of provider batching
tendency. We first present the effects on our main outcomes of length of
stay, 72hr readmission, and healthcare resource utilization (number of
tests).

>>>>

In this section, we examine the causal effects of provider batching
tendency. Our reduced form regression results, as presented in Table
\ref{tab:reducedform}, indicate that the physician's batch tendency is
associated with longer emergency department length of stay (ED LOS), a
greater number of tests ordered, and does not significantly influence
the likelihood of a 72-hour return. However, the relationship between
batch tendency and our outcomes of interest is nuanced by the presence
of testing inclination among physicians.

\begin{table}[htbp]
\centering
\caption{Reduced Form Regression Results with Scaled Coefficients}
\label{tab:reducedform}
\begin{tabular}{p{6cm}ccc}
\toprule
& \multicolumn{1}{c}{\textbf{Log ED LOS}}
& \multicolumn{1}{c}{\textbf{Number of Tests}}
& \multicolumn{1}{c}{\textbf{72hr Return}} \\
& \multicolumn{1}{c}{(1)}
& \multicolumn{1}{c}{(2)}
& \multicolumn{1}{c}{(3)} \\
\midrule
\textit{Model 1} & & & \\
Batch Tendency      & $0.138^{***}$ & $0.270^{***}$ & $-0.004$  \\
                    & $(0.047)$  & $(0.052)$     & $(0.003)$ \\
\\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Patient observables? & Yes & Yes & Yes \\
Hospital volume? & Yes & Yes & Yes \\
\midrule
\midrule
\textit{Model 2} & & & \\
Batch Tendency      & $-0.018$      & $0.019^{**}$ & $0.003$   \\
                    & $(0.071)$     & $(0.007)$     & $(0.003)$ \\
                    &     &     &  \\
Testing Inclination    & $0.595^{**}$  & $0.962^{***}$ & $-0.027^{**}$ \\
                    & $(0.234)$  & $(0.021)$     & $(0.011)$ \\
\\
Seasonality and shift fixed effects? & Yes & Yes & Yes \\
Chief Complaint? & Yes & Yes & Yes \\
Patient observables? & Yes & Yes & Yes \\
Hospital volume? & Yes & Yes & Yes \\
\midrule
Mean number of encounters per physician & 1805.3 &  &  \\
\midrule
\bottomrule
\end{tabular}
\begin{tablenotes}
\small
\item \textit{Estimates of the reduced form are for the baseline sample described in the text. Coefficients on batch tendency and testing inclination are scaled by the difference between batch tendency and testing inclination respectively of the ninetieth and tenth percentile physicians for interpretability. Seasonality shift fixed effects include Year-Month and Hospital-Day of week-Hour of day fixed effects. Chief complaint comes from the cleaned complaint that the patient came in with at the initial encounter. Patient observables include sex dummy, race/ethnicity, and age bins. Robust standard errors are clustered at the physician level.}
\item $*** p < 0.001$, $** p < 0.01$, $* p < 0.05$.
\end{tablenotes}
\end{table}

Similar to our method used to measure physician batch tendency, we use
the physician's residualized leave-out average test rate to get a
measure of a physicians propensity to test (which we call testing
inclination). A key consideration in our analysis is the strong positive
correlation (r = 0.79) between batch tendency and testing inclination.
Such a high degree of correlation raises the concern that physicians
with a higher testing inclination may batch more often because as a
result of ordering more tests. Failure to control for this inclination
could result in biased estimates of the effect of batch tendency. We
note this assumption made on the causal pathway and further explore this
in the section on mediation analysis.

In Model 2, we control for testing inclination and observe notable
changes in the coefficients for batch tendency. The previously
significant positive association with Log ED LOS diminishes, suggesting
that the initial estimate may have been capturing the combined effects
of batching and testing inclination. Estimated coefficients for batch
tendency are scaled by the difference in batch tendency going from the
ninetieth to tenth percentile in physician batch tendency---equal to
15.8 pp---for interpretability, and coefficients for testing inclination
are scaled by the difference in testing inclination going from the
ninetieth to tenth percentile in physician testing inclination---equal
to 34.9 pp.

Assignment to a physician in the top batching decile (relative to one in
the bottom decile) is associated with a 0.2 unit increase in the number
of diagnostic tests being ordered per patient encounter, without any
additional change in the liklihood of a 72 hour return. With an average
of 1,805 patient encounters per physician in our study period, this
translates to an additional 8,667 diagnostic tests with no perceived
benefit in terms of reducing returns or length of stay.

The effects of testing inclination are also striking. We find that a
movement from a physician in the top testing decile (relative to one in
the bottom decile) is associated with 0.6\% change in the length of stay
in the emergency department, a 0.96 unit increase in the number of
diagnostic tests being ordered per patient encounter, and -0.03 pp
change in the probability of a patient returning within 72 hours. Given
the average number of patient encounters per physician in our study
period this translates to an additional 1088 patient hours in LOS,
41,594 diagnostic tests, and 47 avoided 72 hour returns.
